% !TEX root = main.tex
\section{Some omitted proofs}
\begin{proof}[Proof of \Cref{sixseven}]
    We will prove the LHS inequality, the RHS can be proven in the exact same way : \begin{align*}
        \int a(x) d\mu(x) &= \int \frac{a(x)}{b(x)} b(x) d\mu(x) \\
        &\geq \Bigl(\inf_{x \in spt(\mu)} \frac{a(x)}{b(x)}\Bigr) \int b(x) d\mu(x)
    \end{align*}
    The last inequality is implied by the fact that $\forall x\in \mx : b(x) \geq 0$.
\end{proof}
\begin{proof}[Proof of \Cref{gaussianexpec}]
    Let $Z'$ be a standard Gaussian independent of $Z$, then : 
    \begin{equation*}
        \E\Phi(aZ+b) = \P(aZ+b\geq Z') = \P \Bigl(\frac{Z'-aZ}{\sqrt{1+a^2}} \leq \frac{b}{\sqrt{1+a^2}}\Bigr) = \Phi(\frac{b}{\sqrt{1+a^2}})
    \end{equation*}
    The last equality is given by the fact that $\frac{Z'-aZ}{\sqrt{1+a^2}} \sim \mathcal{N}(0,1)$.
\end{proof} 
\begin{proof}[Proof of \Cref{klmixture}]
Let $X = (X_1, \ldots, X_n) \sim \mathcal{N}(0,1)^{\otimes n}$ and let $\phi$ denote the
standard Gaussian density. The mixture $M_n$ has density
\[
    m_n(x) = \frac{1}{2}\prod_{i=1}^{n}\phi(x_i - \omega)
            + \frac{1}{2}\prod_{i=1}^{n}\phi(x_i + \omega).
\]
The likelihood ratio against $\mathcal{N}(0,1)^{\otimes n}$ is
\[
    \frac{m_n(X)}{\prod_i \phi(X_i)}
    \;=\; \frac{1}{2}\bigl(e^{\omega T - n\omega^2/2} + e^{-\omega T - n\omega^2/2}\bigr)
    \;=\; e^{-n\omega^2/2}\cosh(\omega T),
\]
where $T = \sum_{i=1}^n X_i \sim \mathcal{N}(0, n)$ under $\mathcal{N}(0,1)^{\otimes n}$.
Hence
\[
    \mathrm{KL}\bigl(\mathcal{N}(0,1)^{\otimes n} \,\|\, M_n\bigr)
    = -\mathbb{E}_X\!\left[\log \frac{m_n(X)}{\prod_i \phi(X_i)}\right]
    = \frac{n\omega^2}{2} - \mathbb{E}\bigl[\log\cosh(\omega T)\bigr].
\]
Moreover, $\log\cosh(\omega T) \geq \omega^2 T^2/2 - \omega^4 T^4/12$, hence
\[
    \mathbb{E}\bigl[\log\cosh(\omega T)\bigr]
    \;\geq\; \frac{\omega^2}{2}\,\mathbb{E}[T^2] - \frac{\omega^4}{12}\,\mathbb{E}[T^4].
\]
Since $T \sim \mathcal{N}(0,n)$, $\mathbb{E}[T^2] = n$ and $\mathbb{E}[T^4] = 3n^2$.
Therefore
\[
    \mathrm{KL}\bigl(\mathcal{N}(0,1)^{\otimes n} \,\|\, M_n\bigr)
    \;\leq\; \frac{n\omega^2}{2} - \frac{n\omega^2}{2} + \frac{\omega^4 \cdot 3n^2}{12}
    \;=\; \frac{n^2\omega^4}{4},
\]
as claimed.
\end{proof}
\section{Proof of Lipschitz concentration for Gaussians}
Firstly, note that we can restrict ourselves to smooth Lipschitz functions. Explicitely, if $f$ is an $L$-Lipschitz function, one can consider its mollification $f_\epsilon = f * \eta_\epsilon$ where $\eta$ is the standard mollifier.
 Moreover $f_\epsilon$ is smooth and converges uniformly on compact sets to $f$ as $\epsilon \rightarrow 0$. Now, we want to show the sub-Gaussianity of $f(Z)$ where $f : \R^d \rightarrow \R$ is an $L$-Lipschitz smooth function and $Z$ is a standard Gaussian.
This can be done using the Gaussian log-Sobolev inequality, which we will state in the following form : 

Let $\gamma_n$ denote the standard Gaussian measure on $\R^n$, that is, the law
of $X \sim \N(0, I_n)$. We write $\E$ for expectation under $\gamma_n$ and
$\norm{\cdot}_2$ for the Euclidean norm. For a nonnegative measurable function
$h$ with $\E[h]<\infty$ and $\E[h\log h] < \infty$, the \emph{entropy of $h$
relative to $\gamma_n$} is
\[
    \text{Ent}_{\gamma_n}(h) \coloneqq \E[h \log h] - \E[h] \log \E[h],
\]
with the convention $0\log 0 = 0$.

\begin{lemma}[Gaussian logarithmic Sobolev inequality]
\label{lem:lsi}
For every smooth function $g \colon \R^n \to \R$ with $g, \nabla g \in L^2(\gamma_n)$,
\[
    \text{Ent}_{\gamma_n}\!\bigl( g^2 \bigr) \le 2\, \E\!\bigl[\, \norm{\nabla g}_2^2 \,\bigr] .
\]
\end{lemma}


\begin{lemma}[Herbst argument]
\label{lem:herbst}
Let $f \colon \R^n \to \R$ be smooth with $\norm{\nabla f(x)}_2 \le L$ for every
$x \in \R^n$. Then for every $\lambda \in \R$,
\[
    \E\!\Bigl[\, e^{\lambda\,(f(X) - \E f(X))} \,\Bigr] \le \exp\!\left( \frac{\lambda^2 L^2}{2} \right) .
\]
\end{lemma}

\begin{lemma}[Chernoff bound for sub-Gaussian variables]
\label{lem:chernoff}
Let $Y$ be a real-valued random variable satisfying
$\E[e^{\lambda Y}] \le \exp\!\bigl( \lambda^2 \sigma^2 / 2 \bigr)$ for all
$\lambda > 0$. Then for every $t > 0$,
\[
    \P(Y \ge t) \le \exp\!\left( -\frac{t^2}{2\sigma^2} \right) .
\]
\end{lemma}

\begin{theorem}[Gaussian Lipschitz concentration]
\label{thm:main}
Let $X \sim \N(0, I_n)$ and let $f \colon \R^n \to \R$ be $L$-Lipschitz. Then for
every $t > 0$,
\[
    \P\!\bigl( f(X) - \E f(X) \ge t \bigr) \le \exp\!\left( -\frac{t^2}{2L^2} \right) ,
\]
and consequently
\[
    \P\!\bigl( |{f(X) - \E f(X)}|\ge t \bigr) \le 2\exp\!\left( -\frac{t^2}{2L^2} \right) .
\]
The bound is independent of the dimension $n$.
\end{theorem}

\begin{proof}
By the smoothing reduction (i.e. the reduction to smooth Lipschitz functions) it suffices to prove the
inequality under the additional assumption that $f$ is smooth and satisfies
$\norm{\nabla f(x)}_2 \le L$ for every $x \in \R^n$, since the tail probability
in the conclusion is upper semicontinuous with respect to the pointwise
convergence $f_k \to f$.

Define the centered random variable
\[
    Y \coloneqq f(X) - \E f(X) ,
\]
so that $\E Y = 0$. The Herbst argument (Lemma~\ref{lem:herbst}), whose proof
relies on the Gaussian logarithmic Sobolev inequality (Lemma~\ref{lem:lsi}),
asserts that
\[
    \E\!\bigl[\, e^{\lambda Y} \,\bigr] \le \exp\!\left( \frac{\lambda^2 L^2}{2} \right)
    \qquad \text{for every } \lambda \in \R .
\]
In other words, $Y$ is sub-Gaussian with variance proxy $\sigma^2 = L^2$.

Applying the Chernoff bound (Lemma~\ref{lem:chernoff}) with $\sigma^2 = L^2$ then
yields, for every $t > 0$,
\[
    \P\!\bigl( f(X) - \E f(X) \ge t \bigr) = \P(Y \ge t) \le \exp\!\left( -\frac{t^2}{2L^2} \right) ,
\]
which establishes the upper-tail bound.

For the lower tail, observe that $-f$ is also $L$-Lipschitz, and that
$\E(-f)(X) = -\E f(X)$. Applying the upper-tail bound just established to $-f$ in
place of $f$ gives
\[
    \P\!\bigl( f(X) - \E f(X) \le -t \bigr)
    = \P\!\bigl( (-f)(X) - \E(-f)(X) \ge t \bigr)
    \le \exp\!\left( -\frac{t^2}{2L^2} \right) .
\]
Combining the two tail bounds via a union bound,
\[
    \P\!\bigl( |f(X) - \E f(X)| \ge t \bigr) \le 2\exp\!\left( -\frac{t^2}{2L^2} \right) .
\]
None of the steps above involved any dependence on the dimension $n$, so the
estimate is dimension-free.
\end{proof}
